MultiATSM package - General Guidelines

### 2025-03-24 inst/doc/Guidelines.Rmd
version 1.3.0

The *MultiATSM* package offers estimation routines and various outputs
for single and multi-country affine term structure models (ATSMs). All
the frameworks of this package are based on the unspanned economic risk
framework from **(JPS, 2014)**. In essence, these models assume the
absence of arbitrage opportunities, consider a linear state space
representation of the yield curve dynamics and offer a tractable
approach to simultaneously combine the traditional yield curve factors
(spanned factors) along with macroeconomic and financial variables
(unspanned factors).

@@

The *MultiATSM* package includes modified datasets compared to those
originally used in @CandelonMoura2023 (CM, 2023) and @CandelonMoura2024
(CM, 2024) due to limited availability of certain data sources. In some
cases, simulated data replicating the original methodology is employed.
To simplify the exposition, the database of CM (2024) is used throughout
the following sections to demonstrate the package's functions. These
datasets can be accessed using the `LoadData()` function with the
argument *CM_2024*, and similarly, the CM (2023) data with *CM_2023*.

@@

The second dataset focuses on the time series of risk factors. Using the
terminology of JPS (2014), this dataset includes *(i)* country-specific
**spanned factors** (commonly represented by level, slope, and curvature
-- see Section \@ref(S:SpaFac) for an intuitive explanation of these
factors) and *(ii)* a set of country-specific and global **unspanned
factors**, such as economic growth and inflation. Similar to the bond
yield data, these economic indicators must be constructed by the user.

@@

In order to estimate any ATSM of this package, the user needs to specify
several general model inputs, namely:

@@

```{r}
W_type <- 'Sample Mean' # Method to compute the transition matrix
t_First_Wgvar <- "2000" # First year of the sample
t_Last_Wgvar <-  "2015" # Last year of the sample
```

@@

```{r}
GVARlist <- list( VARXtype = "unconstrained", W_type = "Sample Mean",
                  t_First_Wgvar = "2000", t_Last_Wgvar = "2015") 
```

@@

```{r,, eval=FALSE}
data(CM_GlobalMacroFactors)
data(CM_DomMacroFactors)
data(CM_Yields)

@@

ATSMInputs <- InputsForOpt(t0, tF, ModelType, Yields, GlobalMacroVar, DomesticMacroVar,
                         FactorLabels, Economies, DataFreq,CheckInputs = FALSE)
```

@@

In its current form, the *MultiATSM* package allows for the calculation
of two risk compensation measures: term premia and forward premia. In
particular, a $n$-maturity bond yield can be decomposed as the sum of
its expected ($Exp_t^{(n)}$) and term premia ($TP_t^{(n)}$) components.
Technically: $$
y_t^{(n)} = Exp_t^{(n)} + TP_t^{(n)}
$$ In the package's standard form, the expected short rate term is
computed from time $t$ to $t+n$, which represents the bond's maturity:
$Exp_t^{(n)} = \sum_{h=0}^{n} E_t[y_{t+h}^{(1)}]$. Alternatively, one
could perfom such a decomposition at the forward rate level
($f_t^{(n)}$), specifically
$f_t^{(n)} = \sum_{h=m}^{n} E_t[y_{t+h}^{(1)}] + FP_t^{(n)}$ where
$FP_t^{(n)}$ corresponds to the forward premia. In this case, the user
must specify a binary variable set to $1$ if the computation of forward
premia is desired, or $0$ otherwise. If set to $1$, the user must also
provide a two-element numerical vector containing the maturities
corresponding to the starting and ending dates of the bond maturity.
Example:

```{r}
    WishFPremia <- 1 # Wish to estimate the forward premia: YES: 1, NO:0 
    FPmatLim <- c(60,120)
```

@@

If the user intends to generate confidence bands around the point
estimates of the numerical outputs previously described using a
bootstrap procedure, an additional list of inputs is required.
Specifically:

@@

Once bond yields and economic time series data are gathered and model
features are selected, ATSM estimation proceeds in three steps. First,
estimate the country-specific spanned factors for inclusion in the
global ATSM. Next, estimate the parameters governing risk factor
dynamics under the $P-$measure. Finally, optimize the entire selected
ATSM, including the $Q-$measure parameters.

@@

For *N=3*, the spanned factors are traditionally interpreted as level,
slope, and curvature. The nature of such interpretability results from
the features of the PC weight matrix as illustrated below:

@@

To directly obtain the time series of the country-specific spanned
factors, the user can simply use the `Spanned_Factors()` function as
follows:

@@

```{r}
FactorsChina <- RiskFactors[1:7,]
PdynPara <- VAR(FactorsChina, VARtype= "unconstrained")
```

@@

The `GVAR()` function is designed to estimate a first-order GVAR which
is built from country-specific VARX(1,1,1) models. For a comprehensive
discussion of GVAR models, see @ChudikPesaran2016.

@@

The parameters *K0Z* and *K1Z* are estimated as part of the model's
$P-$dynamics and are known in closed form. Likewise, *r0* and *se* have
analytical solutions and can be factored out of the log-likelihood
function. However, the remaining parameters (*K1XQ* and *SSZ*) require
numerical estimation. @LeSingleton2018 provides standardized routines
(integrated into this package) to set effective initial values,
facilitating the optimization process.

@@

## General template

@@

# B.1) SPECIFIC model inputs
#################################### GVAR-based models ##################################################
GVARlist <- list( VARXtype = "unconstrained", W_type = "Sample Mean", t_First_Wgvar = "2005",
                  t_Last_Wgvar = "2019")
# VARXtype: Available options "unconstrained" or "constrained" (VARX)
# W_type: Method to compute the transition matrix. Options:"Time-varying" or "Sample Mean"
# t_First_Wgvar: First year of the sample (transition matrix)
# t_Last_Wgvar:  Last year of the sample (transition matrix)
#################################### JLL-based models ###################################################
JLLlist <- list(DomUnit =  "China")
# DomUnit: name of the economy of the economic system, or "None" for the model "JLL No DomUnit"
###################################### BRW inputs  ######################################################
WishBC <- 0 # Wish to estimate the model with the bias correction method of BRW (2012): #YES: 1, NO:0
BRWlist <- within(list(flag_mean = TRUE, gamma = 0.05, N_iter = 250, B = 50, checkBRW = TRUE,
                       B_check = 1000),  N_burn <- round(N_iter * 0.15))
# flag_mean: TRUE = compute the mean; FALSE = compute the median
# gamma: Adjustment parameter
# N_iter:  Number of iteration to be conserved
# N_burn:  Number of iteration to be discarded
# B: Number of bootstrap samples
# checkBRW: wishes to perform closeness check: TRUE or FALSE
# B_check: If checkBRW is chosen as TRUE, then choose number of bootstrap samples used in the check

@@

# 2) Minor preliminary work: get the sets of factor labels and  a vector of common maturities
FactorLabels <- LabFac(N, DomVar, GlobalVar, Economies, ModelType)
if(any(ModelType == c("GVAR single", "GVAR multi"))){ GVARlist$DataConnectedness <- TradeFlows}

@@

>>>>

The *MultiATSM* package offers estimation routines and various outputs
for single and multi-country affine term structure models (ATSMs). All
the frameworks of this package are based on the unspanned economic risk
framework from @JoslinPriebschSingleton2014  **(JPS, 2014)**. In essence, 
these models assume the absence of arbitrage opportunities, consider a 
linear state space representation of the yield curve dynamics and offer 
a tractable approach to simultaneously combine the traditional yield curve 
factors (spanned factors) along with macroeconomic variables
(unspanned factors).

@@

The *MultiATSM* package includes modified datasets compared to those
originally used in @CandelonMoura2023 (CM, 2023) and @CandelonMoura2024
(CM, 2024) due to limited availability of certain data sources. As such, 
in some cases, simulated data replicating the original methodology is 
employed. To simplify the exposition, the database of CM (2024) is used 
throughout the following sections to demonstrate the package's functions.
These datasets can be accessed using the `LoadData()` function with the
argument *CM_2024*, and similarly, the CM (2023) data with *CM_2023*.

@@

The second dataset focuses on the time series of risk factors. Using the
terminology of JPS (2014), this dataset includes *(i)* country-specific
**spanned factors** (commonly represented by level, slope, and curvature
-- see Section \@ref(S:SpaFac) for an intuitive explanation of these
factors) and *(ii)* a set of country-specific and global **unspanned
factors**, namely macroeconomic variables such as economic growth and 
inflation. Similar to the bond yield data, these economic indicators 
must be constructed by the user.

@@

In order to estimate any ATSM of this package, the user needs to specify
several general inputs, namely:

@@

```{r}
W_type <- 'Sample Mean' # Method to compute the transition matrix
t_First_Wgvar <- "2000" # First year of the sample
t_Last_Wgvar <-  "2015" # Last year of the sample
DataConnectedness <- TradeFlows # Measure of connectedness across countries
```

@@

```{r}
GVARlist <- list( VARXtype = "unconstrained", W_type = "Sample Mean", t_First_Wgvar = "2000",
                   t_Last_Wgvar = "2015", DataConnectedness = TradeFlows) 
```

@@

```{r, eval=FALSE} 
LoadData("CM_2024")

@@

ATSMInputs <- InputsForOpt(t0, tF, ModelType, Yields, GlobalMacroVar, DomesticMacroVar,  
                           FactorLabels, Economies, DataFreq, CheckInputs = FALSE)
```

@@

In its current form, the *MultiATSM* package allows for the calculation
of two risk compensation measures: term premia and forward premia. In
particular, a $n$-maturity bond yield can be decomposed as the sum of
its expected short-rate ($Exp_t^{(n)}$) and term premia 
($TP_t^{(n)}$) components.
Technically: $$
y_t^{(n)} = Exp_t^{(n)} + TP_t^{(n)}
$$ In the package's standard form, the expected short rate term is
computed from time $t$ to $t+n$, which represents the bond's maturity:
$Exp_t^{(n)} = \sum_{h=0}^{n} E_t[y_{t+h}^{(1)}]$. Alternatively, one
could perfom such a decomposition at the forward rate level
($f_t^{(n)}$), specifically
$f_t^{(n)} = \sum_{h=m}^{n} E_t[y_{t+h}^{(1)}] + FP_t^{(n)}$ where
$FP_t^{(n)}$ corresponds to the forward premia. In this case, the user
must specify a binary variable set to $1$ if the computation of forward
premia is desired, or $0$ otherwise. If set to $1$, the user must also
provide a two-element numerical vector containing the maturities
corresponding to the starting and ending dates of the bond maturity.
Example:

```{r}
    WishFPremia <- 1 # Wish to estimate the forward premia: YES: 1, NO:0 
    FPmatLim <- c(60, 120)
```

@@

An additional list of inputs is required if the user intends to 
generate confidence bands around the point estimates of the numerical 
outputs. These results are achieved using a bootstrap procedure.
Specifically, the list of inputs include:

@@

Once bond yields and other (macro)economic time series data are gathered 
and model features are selected, ATSM estimation proceeds in three steps. 
First, the package estimates the country-specific spanned factors for 
inclusion in the global ATSM. Next, it estimates the parameters governing 
risk factor dynamics under the $P-$measure. Finally, it optimizes the 
entire selected ATSM, including the $Q-$measure parameters.

@@

For $N = 3$, the spanned factors are traditionally interpreted as level,
slope, and curvature.  This interpretation arises from the structure of 
the principal component (PC) weight matrix, as shown below :

@@

To directly obtain the time series of the country-specific spanned
factors, the user can use the `Spanned_Factors()` function as
follows:

@@

```{r}
FactorsChina <- RiskFactors[1:7, ]
PdynPara <- VAR(FactorsChina, VARtype= "unconstrained")
```

@@

The `GVAR()` function is designed to estimate a first-order GVAR which
is built from country-specific VARX(1,1,1) models. For a comprehensive
discussion of GVAR models, see, for instance, @ChudikPesaran2016.

@@

The parameters *K0Z* and *K1Z* are estimated as part of the model's
$P-$dynamics and are known in closed form. Likewise, *r0* and *se* have
analytical solutions and can be factored out of the log-likelihood
function. However, the remaining parameters (*K1XQ* and *SSZ*) require
numerical estimation. @LeSingleton2018 provide standardized routines
(integrated into this package) to set effective initial values,
facilitating the optimization process.

@@

## General template {#S:SectionTemplate}

@@

# B.1) SPECIFIC model inputs
#################################### GVAR-based models ##################################################
GVARlist <- list( VARXtype = "unconstrained", W_type = "Sample Mean", t_First_Wgvar = "2005",
                  t_Last_Wgvar = "2019", DataConnectedness <- TradeFlows ) 
# VARXtype: Available options "unconstrained" or "constrained" (VARX)
# W_type: Method to compute the transition matrix. Options:"Time-varying" or "Sample Mean"
# t_First_Wgvar: First year of the sample (transition matrix)
# t_Last_Wgvar:  Last year of the sample (transition matrix)
# DataConnectedness: measure of connectedness across countries
#################################### JLL-based models ###################################################
JLLlist <- list(DomUnit =  "China")
# DomUnit: name of the economy of the economic system, or "None" for the model "JLL No DomUnit"
###################################### BRW inputs  ######################################################
WishBC <- 0 # Wish to estimate the model with the bias correction method of BRW (2012): #YES: 1, NO:0
BRWlist <- within(list(flag_mean = TRUE, gamma = 0.05, N_iter = 250, B = 50, checkBRW = TRUE,
                       B_check = 1000),  N_burn <- round(N_iter * 0.15))
# flag_mean: TRUE = compute the mean; FALSE = compute the median
# gamma: Adjustment parameter
# N_iter:  Number of iteration to be conserved
# N_burn:  Number of iteration to be discarded
# B: Number of bootstrap samples
# checkBRW: wishes to perform closeness check: TRUE or FALSE
# B_check: If checkBRW is chosen as TRUE, then choose number of bootstrap samples used in the check

@@

# 2) Minor preliminary work: get the sets of factor labels and  a vector of common maturities
FactorLabels <- LabFac(N, DomVar, GlobalVar, Economies, ModelType)

@@

## User-Friendly Inspection of Key Elements in the Selected ATSM Specification

Starting from version $1.3.0$, the *MultiATSM* package provides additional ways to concisely inspect key elements of the selected ATSM. In particular, users can now gain an alternative perspective on the general model inputs, optimized model parameters, and root mean square error (RMSE) from the forecast analysis. For clarity, the variable names used hereafter align with those presented in Section \@ref(S:SectionTemplate).

### ATSM inputs
The *ATSMInputs* variable is an object of class *ATSMModelInputs* that includes the $S3$ methods `print()` and `summary()`. The `print()` method provides an overview of both the model inputs and the economic system's general features, while the `summary()` method offers statistical summaries of the risk factors and bond yields.

```{r}
data("Out_Example")
print(Out$ATSMInputs)
```

```{r}
summary(Out$ATSMInputs) 
```

### ATSM parameters
The `Optimization()` function returns an object of class *'ATSMModelOutputs'*, which includes the `summary()` $S3$ method. This method offers a parsimonious summary of key model outputs following 
optimization process, including the maximum log-likelihood value, the intercept and slope coefficients from the model’s cross-sectional dimension, and the count of physical-measure eigenvalues greater than one—an indicator of potential explosive behavior in the model’s time series dimension.

```{r}
summary(Out$ModelParaList) 
```

### Forecasts
The `ForecastYields()` function returns an object of class *'ATSMModelForecast'*, which can be visualized using the `plot()` $S3$ method. This method generates a graphical representation of the RMSEs across different countries and forecast horizons.

```{r}
plot(Out$Forecasts) 
```
